Detection and replication of epistasis influencing transcription in humans

### 2013-09-06 docs/manuscript/manuscript.tex
rebuttal documents sent to peter
work continues on rebuttal

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are encouraging with regards to the detection and replication of epistasis.

@@

\begin{figure}[H]
	\centering
	% \includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression). Columns of GP maps are for each independent dataset. Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}. There is a clear trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}

\begin{figure}[H]
	\centering
	% \includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line. The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}

\begin{figure}[H]
	\centering
	% \includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
	\label{fig:fireworks}
\end{figure}

@@

>>>>

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects (Supplementary Figure \ref{fig:ld_mean}) \cite{Weir2008, Hemani2013}, and for higher power terms of LD $r$ the sampling variance increases (Methods, Supplementary Figure \ref{fig:ld_sd}). Therefore these results are encouraging with regards to the detection and replication of epistasis.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression). Columns of GP maps are for each independent dataset. Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}. There is a clear trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{qqMeta.pdf}
	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions. Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale). The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line. The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets. The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values. Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
	\label{fig:qqMeta}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
	\label{fig:fireworks}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_mean}
	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
	\label{fig:ld_mean}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_sd}
	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As the power term of $r$ increases the sampling variance also increases. Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
	\label{fig:ld_sd}
\end{figure}
\clearpage
